---
title: 'Reproducible Research:  Peer Assessment 1'
output:
  html_document:
    keep_md: yes
  pdf_document: default
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
knitr::opts_chunk$set(warning = FALSE)
knitr::opts_chunk$set(message = FALSE)
```
## Loading and preprocessing the data

```{r}
unzip("activity.zip", exdir = "raw_data")
activityData <- read.csv("raw_data/activity.csv", stringsAsFactors = FALSE)

```
Transforming columns.
```{r}
activityData$date <- as.Date(activityData$date)
library(data.table)
activityData <- as.data.table(activityData)
```


## What is mean total number of steps taken per day?
The sum of all steps taken, disregarding the NA's, is:
```{r}
sum(activityData$steps, na.rm = TRUE)
```

```{r}
stepsByDay <- activityData[, 
                           .(sumDate = sum(steps, na.rm = TRUE)), 
                           by = "date"]
library(ggplot2)
ggplot(data = stepsByDay) + 
  geom_histogram(aes(sumDate)) +
  labs(x = "Steps taken per day")
```

The mean and median of steps taken each day are easily calculed using the new data table:
```{r}
mean(stepsByDay$sumDate)
median(stepsByDay$sumDate)
```

## What is the average daily activity pattern?

```{r}
stepsByInterval <- activityData[, 
                           .(intervalMean = mean(steps, na.rm = TRUE)), 
                           by = "interval"]
ggplot(data = stepsByInterval) +
    geom_line(aes(interval, intervalMean))
```
The interval with the largest average across all days is the following:
```{r}
stepsByInterval[intervalMean == max(stepsByInterval$intervalMean)]
```

## Imputing missing values
The total amount of NA's in the data set can be found in the summary:
```{r}
summary(activityData)
```

```{r}
toImpute <- which(is.na(activityData$steps))
NAstepsIntervals <- activityData[toImpute]$interval
intervalMeans <- sapply(NAstepsIntervals, 
       function(x){
           stepsByInterval[interval == x]$intervalMean
           }
       )
activityDataImputed <- copy(activityData)
activityDataImputed[toImpute]$steps <- intervalMeans
```
The new histogram for the steps taken each day then changes to the following graph:
```{r}
stepsByDayImputed <- activityDataImputed[, 
                           .(sumDate = sum(steps, na.rm = TRUE)), 
                           by = "date"]
ggplot(data = stepsByDayImputed) + 
  geom_histogram(aes(sumDate)) +
  labs(x = "Steps taken per day", title = "Histogram imputed data")
```
The new mean and median of steps taken each day are changed to the following:
```{r}
c(originalMean = mean(stepsByDay$sumDate),
imputedMean = mean(stepsByDayImputed$sumDate))
```
```{r}
c(originalMedian = median(stepsByDay$sumDate),
imputedMedian = median(stepsByDayImputed$sumDate))
```
Both mean and median have increased, meaning the missing data was mostly at very busy intervals with large average steps taken.  
The effect on the total steps taken is the following:
```{r}
c(originalSum = sum(activityData$steps, na.rm = TRUE),
imputedSum = sum(activityDataImputed$steps, na.rm = TRUE))
```
```{r}
sum(activityDataImputed$steps, na.rm = TRUE) / sum(activityData$steps, na.rm = TRUE)
```
A 15% increase opposed to the original data!

## Are there differences in activity patterns between weekdays and weekends?

```{r}
weekendDays <- weekdays(as.Date(c("2016-08-27","2016-08-28")))
activityData[, weekend := weekdays(date) %in% weekendDays]
```
Then a histogram can be made that can distinguish between a weekday or weekend day:
```{r}
stepsByInterval <- activityData[, 
                           .(intervalMean = mean(steps, na.rm = TRUE)), 
                           by = c("weekend", "interval")]
ggplot(data = stepsByInterval) +
    geom_line(aes(interval, intervalMean, col = weekend))
```